---
title: "mono_DE_hack.Rmd"
output: html_document

// 2.20.18
// Want to generate State info with Mono, so that I can export metadata and abuse the findAllMarkers() 
// function in Seurat. 
// Not actually doing any DE gene tests in here...leaving all of that to Seurat
//
// This the big deal right here, bruh
---

Clear the environment
  Careful!!
```{r}
rm(list = ls())
```

Load library
```{r}
library(monocle)
```

Load data frames
```{r}
raw_data <- read.csv('all_NE_raw.csv', header=TRUE)
row.names(raw_data) <- raw_data[,1]
raw_data <- raw_data[ ,-1]

meta_data <- read.csv('all_NE_meta.csv', header=TRUE)
row.names(meta_data) <- meta_data[,1]
meta_data <- meta_data[ ,-1]
```

Lets fuck with our metadata
  Want to change type of 'age' field so that we can get nice plots
```{r}
for(i in 1:length(row.names(meta_data))){
  if (meta_data[i,]$age == '1'){
    meta_data[i,]$age <- "E14"
  }
  if (meta_data[i,]$age == '2'){
    meta_data[i,]$age <- "E15"
  }
  if (meta_data[i,]$age == '3'){
    meta_data[i,]$age <- "E17.5"
  }
  if (meta_data[i,]$age == '4'){
    meta_data[i,]$age <- "E19.5"
  }
   if (meta_data[i,]$age == '5'){
    meta_data[i,]$age <- "PN21"
   }
   if (meta_data[i,]$age == '6'){
    meta_data[i,]$age <- "PN90"
   }
   if (meta_data[i,]$age == '7'){
    meta_data[i,]$age <- "PN120"
  }
}
```

Sanity check
```{r}
dim(raw_data)
dim(meta_data)
```

Create annotated data frame from meta_data
```{r}
meta_data_anno <- new("AnnotatedDataFrame", data = meta_data)
```

Create CellDataSet object
```{r}
cds_NE <- newCellDataSet(cellData = as.matrix(raw_data), phenoData = meta_data_anno, featureData = NULL, expressionFamily = negbinomial.size())
```

Estimate size and dispersion
```{r}
cds_NE <- estimateSizeFactors(cds_NE)
cds_NE <- estimateDispersions(cds_NE)
```


CONSTRUCTING SINGLE CELL TRAJECTORIES


1) Choosing genes that define progress
      Using dpFeature technique
```{r}
cds_NE <- detectGenes(cds_NE, min_expr = 0.1)
fData(cds_NE)$use_for_ordering <- fData(cds_NE)$num_cells_expressed > 0.05 * ncol(cds_NE)
```

Find genes expressed in >10 cells
```{r}
all_NE_expressed_genes <- row.names(subset(fData(cds_NE), num_cells_expressed >= 10))
```

Plot PC variance
```{r}
plot_pc_variance_explained(cds_NE, return_all = F, verbose = T, max_components = 100) 
```

Run dimensionality reduction
```{r}
cds_NE <- reduceDimension(cds_NE, max_components = 2, norm_method = 'log', num_dim = 20, reduction_method = 'tSNE', verbose = T)
```

Cluster cells & plot
```{r}
cds_NE <- clusterCells(cds_NE, verbose = F)
plot_cell_clusters(cds_NE)
```

Refine clustering params
```{r}
plot_rho_delta(cds_NE, rho_threshold = 12, delta_threshold = 30)
```

...and recluster
```{r}
cds_NE_alt <- clusterCells(cds_NE, rho_threshold = 12, delta_threshold = 30, skip_rho_sigma = T, verbose = F)
plot_cell_clusters(cds_NE_alt)
#plot_cell_clusters(cds_NE_alt, color_by = "State")
```

Differential gene expression test
```{r}
clustering_DEG_genes_NE <- differentialGeneTest(cds_NE_alt[all_NE_expressed_genes,], fullModelFormulaStr = '~Cluster', cores = detectCores() - 2)
```

Select top 1000 genes
```{r}
all_NE_ordering_genes <- 
  row.names(clustering_DEG_genes_NE)[order(clustering_DEG_genes_NE$qval)][1:1000]

cds_NE_alt <- setOrderingFilter(cds_NE_alt, ordering_genes = all_NE_ordering_genes)
cds_NE_alt <- reduceDimension(cds_NE_alt, method = 'DDRTree')
cds_NE_alt <- orderCells(cds_NE_alt)
```

...And plot
```{r}
colors <- c("E14" = "steelblue", "E15" = "grey53", "E17.5" = "wheat3", "E19.5" = "darkgoldenrod2", "PN21" = "darkorange2", "PN90" = "firebrick", "PN120" = "red")

plot_cell_trajectory(cds_NE_alt, color_by = "State")
plot_cell_trajectory(cds_NE_alt, color_by = "origin")
plot_cell_trajectory(cds_NE_alt, color_by = "age") + scale_color_manual(values = colors)
plot_cell_trajectory(cds_NE_alt, color_by = "Cluster")
plot_cell_trajectory(cds_NE_alt, color_by = "Pseudotime")
```

Truly dont think this will work
    Holy shit did this just work
      I actually dont believe it
        I think the gods of science just smiled down on me
```{r}
cds_NE_alt_r <- orderCells(cds_NE_alt, root_state = 5)
plot_cell_trajectory(cds_NE_alt_r)

pdf("NE_traj_map_pseudotime_color.pdf", height = 5, width = 10)
plot_cell_trajectory(cds_NE_alt_r, color_by = "Pseudotime")
dev.off()

pdf("NE_traj_map_age_color.pdf", height = 5, width = 10)
plot_cell_trajectory(cds_NE_alt_r, color_by = "age") + scale_color_manual(values = colors)
dev.off()
```

And another plot
```{r}
plot_cell_clusters(cds_NE_alt_r)
plot_cell_clusters(cds_NE_alt_r, color_by = "State")
```

Lets try a new way of plotting these trajectories
```{r}
# most basic case -- this works
plot_complex_cell_trajectory(cds_NE_alt_r, color_by = 'Cluster')
plot_complex_cell_trajectory(cds_NE_alt_r, color_by = 'State')
plot_complex_cell_trajectory(cds_NE_alt_r, color_by = 'age')

# lets try a more complicated one
pdf("allNE_dendro.pdf", width = 10, height = 6)
plot_complex_cell_trajectory(cds_NE_alt_r, root_states = 5, color_by = 'age', cell_size = 0.9, cell_link_size = 0.7) + scale_size(range = c(0.2, 0.2)) + scale_color_manual(values = colors)
dev.off()
```

Can we export metadata? 
  Need to add a 'state' col first
```{r}
meta_data$State<-NA

for(i in 1:length(row.names(meta_data))){
  meta_data$State[i] <- cds_NE_alt_r$State[i]
}

#write.csv(meta_data, "meta_data_mono.csv")
```


////////////////////////////////////////////////////////////////////////////////////////////
////////////////////////////////////////////////////////////////////////////////////////////
////////////////////////////////////////////////////////////////////////////////////////////
////////////////////////////////////////////////////////////////////////////////////////////


BEAM Analysis


Lets see if we can do some branch analysis
    BEAM (Branched Expression Analysis Modeling) finds genes that drive the branching event at the specified branch point.
```{r}
all_BEAM1 <- BEAM(cds_NE_alt_r, branch_point = 1, cores = detectCores() - 2)
all_BEAM1 <- all_BEAM1[order(all_BEAM1$qval),]
```

....and plot
not really working, but should be
```{r}
pdf("BEAM1.pdf", height = 8, width = 8)
plot_genes_branched_heatmap(cds_NE_alt_r[row.names(subset(all_BEAM1, qval < 1e-30)),], branch_point = 1, cluster_rows = T, show_rownames = T)
dev.off()
```

Sorting BEAM genes so that I'm only grabbing the ones with the highest qval that are expressed in >100 cells. Call this 'sub_BEAM'
```{r}
dim(all_BEAM1)

all_BEAM1_order <- all_BEAM1[order(all_BEAM1$qval),]
dim(all_BEAM1_order)

sub_BEAM1 <- all_BEAM1_order[1:100,]
sub_BEAM1 <- sub_BEAM1[row.names(subset(sub_BEAM1, num_cells_expressed > 100)),]
dim(sub_BEAM1)

write.csv(sub_BEAM1, file = "sub_BEAM1.csv")
```

Lets repeat for branch point 2
```{r}
all_BEAM2 <- BEAM(cds_NE_alt_r, branch_point = 2, cores = detectCores() - 2)
all_BEAM2 <- all_BEAM2[order(all_BEAM2$qval),]
```

....and plot
```{r}
pdf("BEAM2.pdf", height = 8, width = 8)
plot_genes_branched_heatmap(cds_NE_alt_r[row.names(subset(all_BEAM2, qval < 1e-15)),], branch_point = 2, cluster_rows = T, show_rownames = T)
dev.off()
```

Sorting BEAM genes so that I'm only grabbing the ones with the highest qval that are expressed in >100 cells. Call this 'sub_BEAM'
```{r}
dim(all_BEAM2)

all_BEAM2_order <- all_BEAM2[order(all_BEAM2$qval),]
dim(all_BEAM2_order)

sub_BEAM2 <- all_BEAM2_order[1:100,]
sub_BEAM2 <- sub_BEAM2[row.names(subset(sub_BEAM2, num_cells_expressed > 100)),]
dim(sub_BEAM2)

write.csv(sub_BEAM2, file = "sub_BEAM2.csv")
```


////////////////////////////////////////////////////////////////////////////////////////////
////////////////////////////////////////////////////////////////////////////////////////////
////////////////////////////////////////////////////////////////////////////////////////////
////////////////////////////////////////////////////////////////////////////////////////////

